partial: An R Package for Creating Partial Dependence Plots

### 2016-08-09 manuscript/greenwell.tex
minor edits

For illustration, we will use the (corrected) Boston housing data which are available from the \CRANpkg{mlbench} package \citep{mlbench-pkg}. These data contain the median value of owner-occupied homes in 506 U.S. census tracts in the Boston area, along with 13 independent variables such as the per capita crime rate by town. We begin by loading the data and omitting unimportant columns.
\begin{example}
data(BostonHousing2, package = "mlbench")  # mlbench must be installed!
boston <- BostonHousing2[, -c(1, 2, 5)]
\end{example}
Next, we fit a random forest to the entire data set with default tuning parameters:
\begin{example}
library(randomForest)
set.seed(101)  # for reproducibility
boston.rf <- randomForest(cmedv ~ ., data = boston, importance = TRUE)
varImpPlot(boston.rf)  # variable importance plot
\end{example}
The model fit is reasonable, with an \dfn{out-of-bag} (pseudo) $R^2$ of 0.89. The variable importance scores are displayed in Figure~\ref{fig:plotmo_vs_partial}. Both plots indicate that the percentage of lower status of the population (\code{lstat}) and the average number of rooms per dwelling (\code{rm}) are highly associated with the median value of owner-occupied homes (\code{cmedv}). The question then arises, "What is the nature of these associations?" To help answer this, we can look at the partial dependence of \code{cmedv} on \code{lstat} and \code{rm}, both individually and together.

@@

\begin{figure}[htbp]
  \centering
  \includegraphics[width=1.0\linewidth]{pd_lstat}
  \caption{Partial dependence of \code{cmedv} on \code{lstat}. \textit{Left}: Default plot. \textit{Right}: Customized plot obtained using the \code{plotPartial} function.}
  \label{fig:pd_lstat}
\end{figure}

@@

The benefit of using \code{partial} is threefold: (1) it is a generic function that can be used for various types of model fits (not just random forests), (2) it will allow for any number of predictors to be used, and (3) it can utilize any of the parallel backends supported by the \CRANpkg{foreach} package \citep{foreach-pkg}; we discuss parallel execution in a later section. For example, the following code chunk could be used to assess the joint effect of \code{lstat} and \code{rm} on \code{cmedv} based on a conditional random forest using various options in \code{plotPartial}. The results are displayed in Figure~\ref{fig:pd_lstat_rm}.
\begin{example}
library(party)
boston.crf <- cforest(cmedv ~ ., data = boston)
pd.lstat.rm <- partial(boston.rf, pred.var = c("lstat", "rm"))
plotPartial(pd.lstat.rm)
plotPartial(pd.lstat.rm, col.regions = colorRampPalette(c("white", "blue")))
plotPartial(pd.lstat.rm, contour = FALSE, zlab = "cmedv", drape = TRUE,
            colorkey = TRUE, screen = list(z = -20, x = -60))
\end{example}
Note that the default color map for contour plots is the Matplotlib "viridis" color map provided through the \CRANpkg{viridis} package \citep{viridis-pkg}.

\begin{widefigure}[htbp]
  \centering
  \includegraphics[width=1.0\linewidth]{pd_lstat_rm}
  \caption{Partial dependence of \code{cmedv} on \code{lstat} and \code{rm}. \textit{Left}: Default plot. \textit{Middle}: Using a different color palette. \textit{Right}: Using a 3-D surface.}
  \label{fig:pd_lstat_rm}
\end{widefigure}

>>>>

For illustration, we will use the (corrected) Boston housing data which are available from the \CRANpkg{mlbench} package \citep{mlbench-pkg}. These data contain the median value of owner-occupied homes in 506 U.S. census tracts in the Boston area, along with 13 independent variables such as the per capita crime rate by town. We begin by loading the data and omitting unimportant columns.
\begin{example}
data(BostonHousing2, package = "mlbench")  # mlbench must be installed!
boston <- BostonHousing2[, -c(1, 2, 5)]
\end{example}
Next, we fit a random forest to the entire data set with default tuning parameters and 500 trees:
\begin{example}
library(randomForest)
set.seed(101)  # for reproducibility
boston.rf <- randomForest(cmedv ~ ., data = boston, importance = TRUE)
varImpPlot(boston.rf)  # variable importance plot
\end{example}
The model fit is reasonable, with an \dfn{out-of-bag} (pseudo) $R^2$ of 0.89. The variable importance scores are displayed in Figure~\ref{fig:plotmo_vs_partial}. Both plots indicate that the percentage of lower status of the population (\code{lstat}) and the average number of rooms per dwelling (\code{rm}) are highly associated with the median value of owner-occupied homes (\code{cmedv}). The question then arises, "What is the nature of these associations?" To help answer this, we can look at the partial dependence of \code{cmedv} on \code{lstat} and \code{rm}, both individually and together.

@@

\begin{figure}[htbp]
  \centering
  \includegraphics[width=1.0\linewidth]{pd_lstat}
  \caption{Partial dependence of \code{cmedv} on \code{lstat} based on a random forest. \textit{Left}: Default plot. \textit{Right}: Customized plot obtained using the \code{plotPartial} function.}
  \label{fig:pd_lstat}
\end{figure}

@@

The benefit of using \code{partial} is threefold: (1) it is a generic function that can be used for various types of model fits (not just random forests), (2) it will allow for any number of predictors to be used, and (3) it can utilize any of the parallel backends supported by the \CRANpkg{foreach} package \citep{foreach-pkg}; we discuss parallel execution in a later section. For example, the following code chunk could be used to assess the joint effect of \code{lstat} and \code{rm} on \code{cmedv} based on a conditional random forest using various options in \code{plotPartial}. The results are displayed in Figure~\ref{fig:pd_lstat_rm}.
\begin{example}
library(party)
boston.crf <- cforest(cmedv ~ ., data = boston)
pd.lstat.rm <- partial(boston.crf, pred.var = c("lstat", "rm"))
plotPartial(pd.lstat.rm)
plotPartial(pd.lstat.rm, col.regions = colorRampPalette(c("white", "blue")))
plotPartial(pd.lstat.rm, contour = FALSE, zlab = "cmedv", drape = TRUE,
            colorkey = TRUE, screen = list(z = -20, x = -60))
\end{example}
Note that the default color map for contour plots is the Matplotlib "viridis" color map provided through the \CRANpkg{viridis} package \citep{viridis-pkg}.

\begin{widefigure}[htbp]
  \centering
  \includegraphics[width=1.0\linewidth]{pd_lstat_rm}
  \caption{Partial dependence of \code{cmedv} on \code{lstat} and \code{rm} based on a conditional random forest. \textit{Left}: Default plot. \textit{Middle}: Using a different color palette. \textit{Right}: Using a 3-D surface.}
  \label{fig:pd_lstat_rm}
\end{widefigure}
